---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# birdnames

<!-- badges: start -->
<!-- badges: end -->

The goal of birdnames is to help with data management for multi-species bird datasets, especially long-term monitoring datasets. All functions in this package operate on bird species names, and provide functionality to move between the different names that might be used for birds. Particularly, this package helps move from raw data with 4-letter alpha-code (banding code) to analyses, reports, etc. with common and scientific names. It also allows filtering by multiple taxonomic levels. 

## Installation

You can install the development version of birdnames from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("scottfjennings/birdnames")
```

## Example workflow using birdnames

Suppose you have a long term monitoring dataset for waterbirds and raptors at a wetland, with the species identity stored as the 4-letter (banding) alpha code. There may be some alpha codes from early in the study that do not reflect changes in bird names (e.g. "OLDS"). There may also be data representing counts of individuals that were identified to group rather than species (e.g. "SCAUP" for Greater and Lesser Scaup). The functions in birdnames can be piped together using magrittr::%>%; in this example we'll build up a workflow one function at a time. 

```{r example, message=FALSE}
library(birdnames)
library(tidyverse)

## Generate basic example dataset
waterbird_data <- data.frame(alpha.code = c("OLDS", "LTDU", "MALL", "WTKI", "GRSC", "SCAUP", "GWTE", "COLO", "WEGR"))
waterbird_data
```

A first step in using this dataset might be to replace outdated species codes with current ones. Note update_alpha() operates on the alpha.code field directly, not on the entire dataframe, so it must be used inside dplyr::mutate().

```{r}
waterbird_data %>% 
  mutate(alpha.code = update_alpha(alpha.code))
```

Next you may want to filter to just the waterbird species (remove White-tailed Kite). bird_taxa_filter combines the tasks of joining the user's dataset to a table with full taxonomic information then filtering based on those taxonomies. The function relies on a reference list of species composed of both the available bird taxonomy compiled in full_bird_list, plus those species, subspecies and species groups that do not exist or do not have complete information in full_bird_list (these additional "custom" species are contained in custom_bird_list). Using combined_bird_list allows proper handling of species groups like "SCAUP".

Note:  Future versions will hopefully allow the user to choose to use full_bird_list, combined_bird_list, or supply their own customized list. I have tried implementing this but there is some aspect of scoping that I don't understand and I can't get it to work.
 
```{r}
waterbird_data %>% 
  mutate(alpha.code = update_alpha(alpha.code)) %>% 
  bird_taxa_filter(keep_taxa = c("Anseriformes", "Gaviiformes", "Podicepiformes"))
```




At this point, the user can proceed with analysis, summary tasks, etc. When it is time to present results, summary tables, etc., species' common names, scientific names, or other taxonomic information can be added with translate_bird_names:  
```{r}
waterbird_data %>% 
  mutate(alpha.code = update_alpha(alpha.code)) %>% 
  bird_taxa_filter(keep_taxa = c("Anseriformes", "Gaviiformes", "Podicepiformes")) %>% 
  mutate(common.name = translate_bird_names(alpha.code, "alpha.code", "common.name"),
         scientific.name = translate_bird_names(alpha.code, "alpha.code", "species"))
```

The provided dataset full_bird_list can be passed to bird_taxa_filter to generate a list of all species meeting a particular taxonomic criteria (i.e. not just those species in the user's own dataset). The user may also want to include species taxonomic number ("species.number") to allow taxonomic sorting of their bird list.

```{r}
data("full_bird_list")
ducks <- full_bird_list %>% 
  select(alpha.code) %>% 
  bird_taxa_filter(keep_taxa = "Anseriformes") %>% 
  mutate(common.name = translate_bird_names(alpha.code, "alpha.code", "common.name"),
         scientific.name = translate_bird_names(alpha.code, "alpha.code", "species"),
         species.number = translate_bird_names(alpha.code, "alpha.code", "species.number")) %>% 
  arrange(species.number)
head(ducks)

``` 


bird_taxa_filer also allows filtering at multiple taxonomic levels. For example, you might wish to isolate data for all waterbirds in the orders Anseriformes, Gaviiformes and Podicepiformes, and also the raptor species in the genera Falco and Accipiter.
```{r}
waterbirds_faclons_accipiters <- full_bird_list %>% 
  select(alpha.code) %>% 
  bird_taxa_filter(keep_taxa = c("Anseriformes", "Gaviiformes", "Podicepiformes", "Falco", "Accipiter")) %>% 
  mutate(common.name = translate_bird_names(alpha.code, "alpha.code", "common.name"),
         scientific.name = translate_bird_names(alpha.code, "alpha.code", "species"),
         order = translate_bird_names(alpha.code, "alpha.code", "order"),
         family = translate_bird_names(alpha.code, "alpha.code", "family"),
         genus = translate_bird_names(alpha.code, "alpha.code", "genus"),
         species.number = translate_bird_names(alpha.code, "alpha.code", "species.number")) %>% view()
```


Finally, many datasets also have a notes field. bird_names_from_text is a helper to find records where bird species are mentioned in the notes field. Note that some 4-letter alpha codes are whole or partial English words, so some non-bird names may slip through this function (e.g. below Chapman's Swift 4-letter code is CHAS so part of chased is retained in bird.notes).

```{r}
data_with_notes <- data.frame(date = c(rep(Sys.Date()-30, 3), rep(Sys.Date()-60, 3), rep(Sys.Date()-90, 3)),
                              alpha.code = rep(c("COLO", "PBGR", "MALL"), 3),
                              notes = c("cheese and crackers for lunch today",
                                        NA,
                                        "also AMWI and teal",
                                        NA,
                                        "chased by peregrine falcon",
                                        NA,
                                        "rain",
                                        NA,
                                        NA))


data_with_notes <- data_with_notes %>% 
  dplyr::mutate(bird.notes = bird_names_from_text(notes))

data_with_notes

```




